---
title: "homework-03"
author: "Iris Carnahan"
date: "2026-05-24"
format: pdf
message: false # use this to make sure messages don't show up
warning: false # use this to make sure warnings don't show up
---

# Set-Up Tasks

```{r}
#| label: reading-in-packages

library(tidyverse) # general use
library(here) # cleaning data frames
library(janitor) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # calculating the data
# reading in `temp-kelp` csv file
temp_kelp <- read.csv(
  here("data", "temp-kelp.csv")
)
# reading in `personal_data_wk8` csv file
my_data <- read.csv(
  here("data", "personal_data_wk8.csv")
)
```

# Problem 1

## a. Appropriate test

To determine the strength of the relationship between these variables, a Pearson's or Spearman rank correlation can be used. Pearson's correlations measure the direction of a linear relationship with continuous variables, while a Spearman rank correlation measures monotonic relationships between any variables (can be ordinal or continuous).

## b. Visualization

```{r}
#| label: correlation-visualization

# base layer: ggplot using `temp_kelp`
ggplot(data = temp_kelp,
       # x-axis: temperature in Celsius
       # y-axis: kelp elongation rate (cm/day)
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # first layer: geom point to represent individual observations
  # change color from default to `red`
  geom_point(color = "red") +
  # relabeling x- and y-axes, adding a title
  labs(x = "Kelp Elongation Rate (cm/day)",
       y = "Temperature (\u00B0C)") +
  # changing theme
  theme_minimal()
```

## c. 

### Check assumptions

```{r}
#| label: QQ-plot-normality
# base layer: ggplot
ggplot(data = temp_kelp, # starting data frame
aes(sample = kelp_elong)) + # y-axis for QQ plot
# first layer: QQ reference line
geom_qq_line(color = "red") + # adding a color
# second layer: QQ plot
geom_qq()
```

```{r}
#| label: linearity-check
# base layer: ggplot using `temp_kelp`
ggplot(data = temp_kelp,
       # x-axis: temperature in Celsius
       # y-axis: kelp elongation rate (cm/day)
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # first layer: geom point to represent individual observations
  # change color from default to `red`
  geom_point(color = "red") +
  # relabeling x- and y-axes, adding a title
  labs(x = "Kelp Elongation Rate (cm/day)",
       y = "Temperature (\u00B0C)") +
  # changing theme
  theme_minimal()
```

```{r}
#| label: residual-vs-leverage-plot
# create model called `assump_check`
# formula: change is kelp elongation rate as a function of temperature
assump_check <- lm(temp_c ~ kelp_elong, data = temp_kelp)  # data frame: `temp_kelp`
# displays residual vs. fitted plot only
plot(assump_check, which = 5)
```

I checked the assumptions of linearity using a scatter plot, that no major outliers are affecting that line using a residuals vs. leverage plot, and for normality using a Q-Q plot. Given each plot, I would say that this relationship is close enough to be considered normal considering the minor deviations from the theoretical distribution, there is a mostly linear relationship shown by the general downward pattern of the scatterplot and assumed normality, and the residuals are within the Cook's distance lines, so there are no major outliers influencing the model.

### Run test

```{r}
#| label: model-equation-components
# display assumption check to generate outputs
summary(assump_check)
```

```{r}
#| label: predicted-values-of-kelp_elong

ggpredict(assump_check,
          terms = "temp_c [14:22 by 2]") |> 
  plot(show_data = TRUE)

```

```{r}
#| label: correlation-strength
# running test to determine strength of correlation
cor.test(temp_kelp$temp_c, temp_kelp$kelp_elong, method = "pearson")

```



## d.

## e. 

## f.

# 2. 

## a.

## b.

# 3. 

## a.

## b.

## c.

## d.

## e.

# 4.

## a.

## b.

## c.

## d.
