---
title: "homework-03"
author: "Iris Carnahan"
date: "2026-05-24"
format: pdf
message: false # use this to make sure messages don't show up
warning: false # use this to make sure warnings don't show up
---

# Set-Up Tasks

```{r}
#| label: reading-in-packages

library(tidyverse) # general use
library(here) # cleaning data frames
library(janitor) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # calculating the data
# reading in `temp-kelp` csv file
temp_kelp <- read.csv(
  here("data", "temp-kelp.csv")
)
# reading in `personal_data_wk8` csv file
my_data <- read.csv(
  here("data", "my_es_data.csv")
)
# reading in `bike_to_class` csv file
bike_to_class <- read.csv(
  here("data", "bike_to_class.csv")
)
```

# Problem 1

## a. Appropriate test

To determine the strength of the relationship between these variables, a Pearson's or Spearman rank correlation can be used. Pearson's correlations measure the direction of a linear relationship with continuous variables, while a Spearman rank correlation measures monotonic relationships between any variables (can be ordinal or continuous).

## b. Visualization

```{r}
#| label: correlation-visualization

# base layer: ggplot using `temp_kelp`
ggplot(data = temp_kelp,
       # x-axis: temperature in Celsius
       # y-axis: kelp elongation rate (cm/day)
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # first layer: geom point to represent individual observations
  # change color from default to `red`
  geom_point(color = "red") +
  # relabeling x- and y-axes, adding a title
  labs(x = "Kelp Elongation Rate (cm/day)",
       y = "Temperature (\u00B0C)") +
  # changing theme
  theme_minimal()
```

## c. Assumption checks

### Check assumptions

```{r}
#| label: QQ-plot-normality
# base layer: ggplot
ggplot(data = temp_kelp, # starting data frame
aes(sample = kelp_elong)) + # y-axis for QQ plot
# first layer: QQ reference line
geom_qq_line(color = "red") + # adding a color
# second layer: QQ plot
geom_qq()
```

```{r}
#| label: linearity-check
# base layer: ggplot using `temp_kelp`
ggplot(data = temp_kelp,
       # x-axis: temperature in Celsius
       # y-axis: kelp elongation rate (cm/day)
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # first layer: geom point to represent individual observations
  # change color from default to `red`
  geom_point(color = "red") +
  # relabeling x- and y-axes, adding a title
  labs(x = "Kelp Elongation Rate (cm/day)",
       y = "Temperature (\u00B0C)") +
  # changing theme
  theme_minimal()
```

```{r}
#| label: residual-vs-leverage-plot
# create model called `assump_check`
# formula: change is kelp elongation rate as a function of temperature
assump_check <- lm(kelp_elong ~ temp_c, data = temp_kelp)  # data frame: `temp_kelp`
# displays residual vs. fitted plot only
plot(assump_check, which = 5)
```

I checked the assumptions of linearity using a scatter plot, that no major outliers are affecting that line using a residuals vs. leverage plot, and for normality using a Q-Q plot. Given each plot, I would say that this relationship is close enough to be considered normal considering the minor deviations from the theoretical distribution, there is a mostly linear relationship shown by the general downward pattern of the scatterplot and assumed normality, and the residuals are within the Cook's distance lines, so there are no major outliers influencing the model.

### Run test

```{r}
#| label: model-equation-components
# display assumption check to generate outputs
summary(assump_check)
```

```{r}
#| label: predicted-values-of-kelp_elong
# generating the adjusted reductions for response variable
ggpredict(assump_check,
          # selecting a range from 14 to 22 in increments of 2
          terms = "temp_c [14:22 by=2]") |> 
  plot(show_data = TRUE)
```

```{r}
#| label: correlation-strength-p
# running test to determine strength of correlation
cor.test(temp_kelp$temp_c, temp_kelp$kelp_elong, method = "pearson") # pearson's method
```

## d. Communication results

To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used a Pearson's correlation because the variables were continuous, mostly linear, and normally distributed, enabling me to use this measurement.

We found a strong negative relationship between temperature and giant kelp frond elongation rate (Pearson's r = -0.7, t(30) = -5.2, p < 0.01, $\alpha$ = 0.05), where temperature significantly predicted kep elongation rate. From our model coeffiecient estimates, we found that for each 1 degree Celsius increase in temperature, kelp elongation rate decreases 0.43 ± 0.08 cm/day.

## e. How to apply test results

Given the results of the Pearson's correlation test performed, there is generally a negative relationship between temperature and kelp elongation rate. This means that as temperature increases, the elongation and growth rate of these kelp fronds are decreasing 0.43 ± 0.08 cm/day for every degree Celsius increase. If the team desires to maximize kelp frond elongation, maintaining a water temperature around 14°C would maintain a higher elongation rate overall, meaning colder ocean environments are more productive.

## f. Double check

```{r}
#| label: correlation-strength-s
# running test to determine strength of correlation
cor.test(temp_kelp$temp_c, temp_kelp$kelp_elong, method = "spearman") # spearman method
```

Using a Pearson's versus a Spearman correlation would have made no difference, as in both tests there is a strong negative relationship temperature and kelp elongation rate and the null hypothesis would be rejected (Spearman $\rho$ = -0.7, S = 9216.1, p < 0.01, $\alpha$ = 0.05). While the outputs mean the same thing, the test components differ since a Pearson's correlation test outputs correlation coefficient (r), a one-sample t-test comparing r to 0, and a 95% confidence interval, while a Spearman correlation test outputs the correlation coefficient ($/rho$), the sum of all squared rank differences (S), and no confidence intervals. Additionally, within the data, there are tied ranks which leads the p-value commuted to be approximate instead of exact for a Spearman correlation.

# Problem 2

## a. Updating my data

```{r}
#| label: cleaning-my-data
# creating new data frame using `my_data`
my_hist <- my_data |> 
  # cleaning column names so that only lower case letters and underscores are used
  clean_names()
# reordering levels according to the day of the week in chronological order
my_hist$days <- factor(my_hist$days,
       levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))

```


```{r}
#| label: updated-personal-data-visuals

#base layer: ggplot
ggplot(data = my_hist,
       # x-axis should be the day of the week and y-axis should be the time spent biking
       # fill column color to differentiate days
       mapping = aes(x = days,
                     y = time_spent_biking_min,
                     fill = days)) +
  # first layer: bar chart with height proportional to the values of the data
  geom_col(width = 0.5,
           # color border line of bar black
           color = "black",
           # fill bin color pink
           fill = "pink") +
  # relabeling x- and y-axes, adding title
  labs(title = "Comparison of Total Time Spent Biking Throughout the Week",
       x = "Day of the Week",
       y = "Time Biking on Campus (min)") +
  # changing theme of histogram
  theme_minimal() +
  # removing legend
  theme(legend.position = "none")
```

```{r}
#| label: biking-classes-jitter

```



## b. Creating captions

# 3. 

## a.

## b.

## c.

## d.

## e.

# 4.

## a.

## b.

## c.

## d.
