---
title: "homework-03"
author: "Iris Carnahan"
date: "2026-05-24"
format: pdf
---

# Set-Up Tasks

```{r}
#| label: reading-in-packages

library(tidyverse) # general use
library(here) # cleaning data frames
library(janitor) # file/folder organization
library(readxl) # reading .xlsx files

temp_kelp <- read.csv(
  here("data", "temp-kelp.csv")
)

my_data <- read.csv(
  here("data", "personal_data_wk8.csv")
)
```

# Problem 1

## a. Appropriate test

To determine the strength of the relationship between these variables, Pearson's or Spearman's correlations can be used. Pearson's correlations measure that direction of a linear relationship with continuous variables, while a Spearman's correlation measures monotonic relationships between any variables (ordinal or continuous).

## b. Visualization

```{r}
#| label: correlation-visualization


ggplot(data = temp_kelp,
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  geom_point(color = "red") +
  labs(x = "Kelp Elongation Rate (cm/day)",
       y = "Temperature (\u00B0C)") +
  theme_minimal()
```

## c. 

### Check assumptions

### Run test

```{r}
#| label: QQ-plot
# base layer: ggplot
ggplot(data = temp_kelp, # starting data frame
aes(sample = kelp_elong)) + # y-axis for QQ plot
# first layer: QQ reference line
geom_qq_line(color = "red") + # adding a color
# second layer: QQ plot
geom_qq()
```





